---
title: "final_project.rmd"
author: "Emma-Marie Vitskov Jørgensen"
date: "2022-11-07, updated `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Some scholars
For my final project I will investigate if there has been a secularisation tendency in danish funerals. My research question: "What characterizes the development in the number of civil funerals in Denmark from 2006-2021, and does the development reflect the number of deregistrations from the church?"

I will be analyzing a data sets from Danmarks Statistik (https://www.dst.dk/da/). When I downloaded the data sets from the website they were downloaded as an excel file, and I converted it into a .csv format before loading them into this script. 
The data sets can be found in the "data" repository, and information about the three data sets and the variables from this rmarkdown can be found under "Required Metadata" in the project report.


### Loading packages
First step is to load the packages needed for my data processing and analysis.

```{r load packages}
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)

```

### Loading messy data sets
 
Now I'm able to load the raw and messy data from Danmarks Statistik into RMarkdown using `read_csv()` function:

```{r load data}

setwd('..')
funerals_messy <- read_csv("./data/funarals_messy.csv")

#checking the data type
class(funerals_messy) #it's a tibble

```

## Inspect the `funerals_messy` data

I will inspect all the data sets to check if it is clean and ready for analysis

```{r inspect funerals}
#inspecting the data
head(funerals_messy)
tail(funerals_messy)

#viewing the data set
view(funerals_messy)

```

There is several things about `funerals_messy` which I would like to change before the analysis!:
  - All the information is placed in one column, and have to be divided into several columns. 
  - Turn the years from row 2 into the names of column 2-16 and afterwards remove row 2.
  - Change the name of column 1.
  - Remove row 1 which only contains a note about the unit of the data (which is number) and a lot of empty cells (NA). 
  - Remove row 10 which is empty
  - Remove row 11 which is mostly empty except from a note about the source of the data (the danish church ministry). 
  - The values in the first column (the different types of funerals) are quite long and hard to work with. Therefore I will change the names into some shorter and nicer names. I will also translate them from danish to English. 
  -  change data type of the numbers into numeric data, if it isn't already 


## Cleaning the `funerals_messy` data

I start by turning the data into a data frame:

```{r make data frame}
#turn into data frame
funerals_messy <- as.data.frame(funerals_messy)
class(funerals_messy)
```

I start cleaning the `funerals_messy`data set by turning the years into column names.

```{r change column names}
#Turn row 2 into columns names
#checking that 2 is the right row number
funerals_messy[2,] #it is!

#turning row 2 into the column names
colnames(funerals_messy) <- funerals_messy[2,]
head(funerals_messy)

#separating the different years by the ; into individual columns
funerals_messy[c(' ','2006', '2007', '2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018', '2019', '2020', '2021')] <- str_split_fixed(funerals_messy$`;2006;2007;2008;2009;2010;2011;2012;2013;2014;2015;2016;2017;2018;2019;2020;2021`, ';', 17)

head(funerals_messy)# the years are now column names!

```

After viewing the data frame now, I see that every single piece of data has been separated into individual columns under the right year. But column one still contains all the information from all years bundled up in one column. This column is now needless and I will remove it:

```{r remove column}

#removing column 1
funerals_LessColumns <- funerals_messy[,-1]

funerals_LessColumns #1st column is now gone

```

The new column 1 lacks a name:

```{r name column}

#changing name of column 1
colnames(funerals_LessColumns)[1] <- "funeral_type"

#checking if name is changed
head(funerals_LessColumns)
```

Now I remove row 2 which contain the years, row 1 which contain a comment on the data unit, row 10 which is empty and row 11 which contains a comment in source.

```{r remove rows}
#Removing 1st, 2nd, 10th and 11th row
funerals_LessRows <- funerals_LessColumns[-c(1,2,10,11),]

#Checking if rows have been removed
head(funerals_LessRows)
tail(funerals_LessRows)
#The rows have successfully been removed.


```

Now the data set looks quite nice. The final touches will be to change the values in column 1 into something shorter and something english:

```{r rename values}
#changing values

funerals_LessRows[funerals_LessRows == "Dødsfald registreret"] <- "Registered_Deaths"
funerals_LessRows[funerals_LessRows == "Begravet/bisat under medvirken af pastoratets præster fra sognets kirke"] <- "PastoralPriest_ParishChurch"
funerals_LessRows[funerals_LessRows == "Begravet/bisat under medvirken af pastoratets præster andetsteds end fra sognets kirker"] <- "PastoralPriest_OtherChurch"
funerals_LessRows[funerals_LessRows == "Begravet/bisat under medvirken af andre præster i folkekirken fra sognets kirker"] <- "OtherPriest_ParishChurch"
funerals_LessRows[funerals_LessRows == "Begravet/bisat under medvirken af andre præster i folkekirken andetsteds end sognets kirke"] <- "OtherPriest_OtherChurch"
funerals_LessRows[funerals_LessRows == "Begravet/bisat under medvirken af præster fra frimenigheder/andre trossamfund"] <- "FreeChurch_OtherReligion"
funerals_LessRows[funerals_LessRows == "Begravet/bisat uden gejstlig medvirken"] <- "Civil_Funeral"

#Checking if the values have changed
head(funerals_LessRows) # they have!
tail(funerals_LessRows)

```
Finally, I would like to check the data type to see if the years and the numbers are of the numeric data type, and convert them into numeric data if the aren't:

```{r change data type}
 #checking data type of the numbers in the 2006 column
class(funerals_LessRows$`2006`) # the data is characters, but I would like it to be numeric

#changing data types

funerals_LessRows$`2006` <- as.numeric(funerals_LessRows$`2006`)
funerals_LessRows$`2007` <- as.numeric(funerals_LessRows$`2007`)
funerals_LessRows$`2008` <- as.numeric(funerals_LessRows$`2008`)
funerals_LessRows$`2009` <- as.numeric(funerals_LessRows$`2009`)
funerals_LessRows$`2010` <- as.numeric(funerals_LessRows$`2010`)
funerals_LessRows$`2011` <- as.numeric(funerals_LessRows$`2011`)
funerals_LessRows$`2012` <- as.numeric(funerals_LessRows$`2012`)
funerals_LessRows$`2013` <- as.numeric(funerals_LessRows$`2013`)
funerals_LessRows$`2014` <- as.numeric(funerals_LessRows$`2014`)
funerals_LessRows$`2015` <- as.numeric(funerals_LessRows$`2015`)
funerals_LessRows$`2016` <- as.numeric(funerals_LessRows$`2016`)
funerals_LessRows$`2017` <- as.numeric(funerals_LessRows$`2017`)
funerals_LessRows$`2018` <- as.numeric(funerals_LessRows$`2018`)
funerals_LessRows$`2019` <- as.numeric(funerals_LessRows$`2019`)
funerals_LessRows$`2020` <- as.numeric(funerals_LessRows$`2020`)
funerals_LessRows$`2021` <- as.numeric(funerals_LessRows$`2021`)

#just checking a number in 2006 and 2017 to see if the data type has actually changed
class(funerals_LessRows[1,2])
class(funerals_LessRows[1,17]) # the data are now numeric

```


Now that the data set is clean and ready for analysis I'll export the clean data into a csv file called `funerals_clean.csv`:

```{r save clean funerals}
getwd() #get my current position
setwd('..') #changing position

readr::write_csv(funerals_LessRows,"data_output/funerals_clean.csv")# creates a new csv

```

## Filter and analyse the data

To be able to do some visualization with percent I will create a data frame were the `Registered_Deaths` row is removed from the funerals_LessRows and a row with the total sum of funerals for each year is added:

```{r remove row}
 #removing the first row
funerals_noRegDeaths <- funerals_LessRows[-1,]

```

Summing the number of funerals for each year:

```{r sum funerals}
funerals_total <- funerals_noRegDeaths %>%
  bind_rows(summarise_all(., ~if(is.numeric(.)) sum(.) else "Total_Funerals"))

#checking if the row has been created. 
funerals_total

```

I will save the data frame with the total amount of funerals as a separate csv called `funerals_total.csv`:

```{r save total funerals data frame}
setwd('..') #changing position
readr::write_csv(funerals_total,"data_output/funerals_total.csv")# creates a new csv

```

I will create a data frame only containing the religious funerals and the sum of those. This means I will remove the `Registered_Deaths` row and the `Civil_Funeral` row from the `funerals_LessRows` data and then sum the rest of the rows:

```{r remove row 1 and 7}
funerals_religious <- funerals_LessRows[-c(1,7),]

funerals_religious # checking that rows are removed

```


Summing the number of religious funerals for each year:

```{r sum religious funerals}
funerals_rel_total <- funerals_religious %>%
  bind_rows(summarise_all(., ~if(is.numeric(.)) sum(.) else "total_rel_funerals"))

#checking if the row has been created. 
funerals_rel_total

```
Saving the data frame with the total amount of religious funerals in a csv called `funerals_Religious.csv`:

```{r save total funerals data frame}
setwd('..') #changing position
readr::write_csv(funerals_rel_total,"data_output/funerals_religious.csv")# creates a new csv

```


Counting the total amount of funerals held by the danish national church (that means all religious funerals minus the number in the FreeChurch_OtherReligion) for each year: 

```{r sum national church funerals}

funerals_nationalchurch <- funerals_noRegDeaths [1:4, ]

funerals_nationalchurch # checking if the right rows have been chosen. 


```

Summing the number of funerals in the danish national church for each year:

```{r sum religious funerals}
funerals_nationalchurch_total <- funerals_nationalchurch %>%
  bind_rows(summarise_all(., ~if(is.numeric(.)) sum(.) else "total_natchurch_funerals"))

#checking if the row has been created. 
funerals_nationalchurch_total

```


Saving the data frame with the total amount of funerals in the danish national church in a csv called `funerals_NationalChurch.csv`:

```{r save total funerals data frame}
setwd('..') #changing position
readr::write_csv(funerals_nationalchurch_total,"data_output/funerals_nationalchurch.csv")# creates a new csv

```

Save the total amount of funerals, total amount of religious funerals, total amount of funerals in danish national church and total number of civil funerals in the same data frame:

```{r data frame all total values}
#Adding rows from other data sets
total_all_types <- funerals_total[c(6),]
total_all_types[nrow(total_all_types) + 1,] <- funerals_rel_total[5,]
total_all_types[nrow(total_all_types) + 1,] <- funerals_rel_total[6,]
total_all_types[nrow(total_data) + 1,] <- funerals_nationalchurch_total[5,]

#Checking if everything looks right
total_all_types
```
There has been added an empty row, which I will remove:
```{r remove rows 2}
total_all_types2 <- total_all_types[-c(4,5),]

total_all_types2 #the rows have been deleted.
```

Saving the data frame with the total amount of the different funeral types in a csv called `total_alltypes.csv`:

```{r save total funerals data frame}
setwd('..') #changing position
readr::write_csv(total_all_types2,"data_output/total_alltypes.csv")# creates a new csv

```

Now I got the data I need to make the visualizations I have in mind.

#### Vizualisation - Let's see what the funeral trends are

I would like to make a visualization that compares the amount of civil funerals with the amount of religious funerals.I choose a line chard to compare the funeral types from 2006-2021. To be able to do so with the ggplot 2 packages, I turn the data frames into a long format, because that works best for ggplot2:


```{r long format}

#making data frames into long format

total_all_types_long <- total_all_types2 %>% 
  gather(year, number, -funeral_type)

total_all_types_long

```

Now I'm able to make the chard:

```{r line chard}
funerals_plot <- ggplot(total_all_types_long,aes(x=`year`, y=`number`, color=funeral_type)) +
  geom_line() +
  geom_point() +
  labs(title="Danish funerals 2006-2021",x="Year",y="Number of funerals", color="Funeral types",caption = "Figure 1: Long term increase in civil funerals and in funerals from other religions since 2006, but short term decline in civil funerals since 2018.") +
  scale_color_manual(labels = c("civil", "free church and other religion", "national church", "total religious")) +
  scale_color_brewer(palette = "Paired") +
  theme(legend.position="bottom")

funerals_plot # show chart
  
```
From the line chard I can see some trends:

Long term trends: 
- small increase in civil funerals and in funerals in religions and churches not part of the danish national church. 
- Small decline in funerals in the national church. 
Short term trends:
- small decline in civil funerals since 2018-2021. In the same period both funerals in national church and in other religions/free churches are increased.
- small increase in both civil funerals and funerals in national church from 2014-2018, which might just be due to a general rise in the number of funerals. 


Save the chart as PNG in fug_output repository:

```{r save chart 1}
setwd('..') #set the current position

png(file="fig_output/funerals_linechart.png",
width=600, height=350)
funerals_plot 
dev.off()

```

For the next chart I want to know what percentage of the total funerals are civil funerals, funerals in the national church and funerals in other religions/free churches:

```{r percentage dataframe}
#calculating civil funerals percentage
civil_percent <- total_all_types2[1,2:17]/Total_Funerals[1,2:17]*100
natchurch_percent <- total_all_types2[4,2:17]/Total_Funerals[1,2:17]*100
otherrel_percent <- total_all_types2[2,2:17]/Total_Funerals[1,2:17]*100

civil_1digit <- round(civil_percent, digits=1)
natchurch_1digit <- round(natchurch_percent, digits = 1)
otherrel_1digit <- round(otherrel_percent, digits = 1)

```


Creating three separate data frames with the percentage from above and a column containing the funeral type, and making the data frames into long format: 
```{r create data frame}
#create data frame with percent and the funeral type
funeral_type = c("Civil_Funeral")
civil_percent_df <- cbind(funeral_type, civil_1digit)
#Turn into long format
civil_percent_long <- civil_percent_df %>% 
  gather(year, percent, -funeral_type)


#create data frame with percent and the funeral type
funeral_type = c("natchurch_funerals")
natchurch_percent_df <- cbind(funeral_type, natchurch_1digit)
#Turn into long format
natchurch_percent_long <- natchurch_percent_df %>% 
  gather(year, percent, -funeral_type)


#create data frame with percent and the funeral type
funeral_type = c("FreeChurch_OtherReligion")
otherrel_percent_df <- cbind(funeral_type, otherrel_1digit)
#Turn into long format
otherrel_percent_long <- otherrel_percent_df %>% 
  gather(year, percent, -funeral_type)
```


Combining the three data frames with percentages into one, which I can then use for visualization:

```{r combined data frame percent}
#Combine data frames

funerals_percent <- rbind(civil_percent_long,natchurch_percent_long, otherrel_percent_long)

```

Saving the data frame with the  percentages of the different funeral types in a csv called `funerals_percent.csv`:

```{r save total funerals data frame}
setwd('..') #changing position
readr::write_csv(funerals_percent,"data_output/funerals_percent.csv")# creates a new csv

```


Creating a chard showing the percentage of civil and religious funerals for each year:
```{r bar chart}

percent_chart <- ggplot(funerals_percent,aes(x=`year`, y=`percent`, fill=funeral_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = `percent`), colour = "White", size = 4,
    vjust = 1.5, position = position_dodge(.0)) +
   labs(title="Danish funerals 2006-2021",x="Year",y="Percent of total amount of funerals", color="Funeral types",caption = "Figure 1: the percentage of civil funerals and funerals in other religions/free churches have increased since 2006.") +
  scale_fill_brewer(type = "qual", palette = "Paired") +
  scale_color_manual(labels = c("civil", "free church and other religion", "national church")) +
  theme(legend.position="bottom")



percent_chart #show chart
  
```

Save the chart as PNG in fug_output repository:

```{r save chart 1}
setwd('..') #set the current position

png(file="fig_output/percent_barchart.png",
width=600, height=350)
percent_chart 
dev.off()

```


## Conclusion

- Udregn procentdelen af civile begravelser ud af alle religiøse og vis på bjælke-diagram (hvor hver bjælke er 100% og der er en bjælke for hvert år)
- vis antallet af civile, kristne og andre religiøse begravelser i samme chard (tre linjer i spaghettidiagram)
- besvar: hvis stigning i civile - er der så fald af religiøse generelt, eller blot fald af kristne/folkekirkelige? --> er religion på tilbagegand, eller er det bare folkekirken?

