---
title: 'Homework 3: start with R'
author: "Emma-Marie Vitskov Jørgensen"
date: "2022-12-13, updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package}
library(dplyr)
```

# TASK 1

At first I create a vector called "rooms" and assign it with the values given in the homework assignment

```{r vector}
rooms <- c(1, 2, 4, 5, 1, 3, 1, NA, 3, 1, 3, 2, 1, NA, 1, 8, 3, 1, 4, NA, 1, 3, 1, 2, 1, 7, 1, 9, 3, NA)
```


Now I want to remove the NA from the vector

```{r remove NA}
rooms_nona <- rooms[!is.na(rooms)]
rooms_nona
```

Then I create a new vector only containing the elements which is greater than 2:
```{r new vector}
list(rooms_nona[rooms_nona>2])
rooms_above_2 <- rooms_nona[rooms_nona>2]
rooms_above_2
```

Now I want R to count the number of elements greater than 2:
```{r sum}
sum(rooms_above_2)
```

The sum of the elements grater than 2 is 55.

# TASK 2: what type of data is the "rooms" vector?

Checking the data type of the "rooms" vector:
```{r}
is.numeric(rooms) 
```

The vector contains numeric data.

# TASK 3

Installing "tidyverse" package:
```{r packages 2}
install.packages("tidyverse")
library(tidyverse)
```
Downloading the SAFI_clean.csv file into RStudio:
```{r download SAFI}
download.file("https://ndownloader.figshare.com/files/11492171",
              "SAFI_clean.csv", mode = "wb")
interviews <- read_csv("SAFI_clean.csv")
```




# Task 4

Installing tidyverse (to be able to make tibbles)
```{r packages 3}
install.packages("tidyverse")
library(tidyverse)
```

Downloading the monarchs dataset
```{r}
monarchs <- read_csv("data/monarchs.csv")

monarchs #viewing data
class(monarchs)
```

Exchanging the NULLs with NAs (so RStudion interprets it as missing data)
```{r NULL to NA}
monarchs[monarchs == "NULL"] <- NA

monarchs #The NULLs are now NAs
```

Checking the data type of the `years_ruled` column in the data set, to see if its numeric:
```{r}
class(monarchs$years_ruled)
```

The data type is character and not numeric, and therefore I have to change the data type into numeric. But first I filter the NAs away:

```{r remove NA 2}
monarchs_nona <- monarchs %>% filter(years_ruled != "NA")

```

Changing the data in the years_ruled column into numeric data
```{r change data type}
monarchs_nona$years_ruled <- as.numeric(monarchs_nona$years_ruled)
class(monarchs_nona$years_ruled)
```

The data in the `years_ruled` column is now numeric, and I can calculate the mean and median

Calculating the mean duration of rule over time:
```{r mean}
mean(monarchs_nona$years_ruled)
```


Calculating the median:
```{r median}
median(monarchs_nona$years_ruled)
```

The mean is 19.13 years and the median is 16.5 years. It has to be noted, that the reign time of Margrethe 2. isn't a part of the calculations, because her reighn period hasn't ended yet. 

Sorting the data set to see the three largest values (the three longest reigns)
```{r}
sort(monarchs_nona$years_ruled)
```

The three longest periods of reign is 43, 50 and 60 years

Finding out which rows the three longest reigns belongs to:
```{r find rows}
which(monarchs == 60, arr.ind=TRUE)
which(monarchs == 50, arr.ind=TRUE)
which(monarchs == 43, arr.ind=TRUE)
```

The three monarchs ruling the longest can be found in row 40, 54 and 50

Finding the names of the monarchs who have the longest reign periods
```{r}
print(monarchs[40,])
print(monarchs[54,])
print(monarchs[50,])
```

The monarchs ruling the longest are Christian 4th, Margrethe 2nd and Christian 9th. 

How many days did they rule?
```{r days of ruling}
#Christian 4th (row 39)
60*365
# Margrethe 2nd (row 53)
50*365
# Christian 9th (row 49)
43*365
```

Christian 9th has ruled for 21900 days, Margrethe 2nd for 18250 days and Christian 9th for 15695 days. 
NB: I multiply by 365, because this is the average length of a year, but the number of days might be slightly incorrect, because of leap years etc.