---
title: "Homework 6: Practicing functions with Gapminder"
author: "Emma-Marie"
date: "2022-10-31, updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1
I need to define a defensive function that calculates the Gross Domestic Product of a nation from the data available in the gapminder dataset. 
I started out by downloading the gapminder dataset and named it `gapminder_data`. 

```{r load data}
library(readr)
gapminder_data <- read_csv("./data/gapminder_data.csv")

show(gapminder_data)

```
Because I uploaded the data with the read_csv() function, it was uploaded as a tibble. This means, that I can't calculate the mean, which is necessary for me to do in task two. Therefore I convert the data into a data frame called `gapminder` using the as.data.frame() function: 

```{r change data type}
class(gapminder_data) #checking data type: its a tible
#changing the data type into a data frame
gapminder <- as.data.frame(gapminder_data)

class(gapminder) #checking the data type of "gapminder" it's a data frame
```

I define a defensive function that calculates the Gross Domestic Product of a given year or nation from the data set: 
```{r function}
# Defining the defensive function
calcGDP <- function(dat, year=NULL, country=NULL) {
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
  gdp <- dat$pop * dat$gdpPercap
  
  new <- cbind(dat, gdp=gdp)
  return(new)
}
```

I calculat the GDP of Denmark in year 1967, 1977, 1987, 1997, 2007 and 2017 using the function above, by typing in “Denmark” and the given years in the function. 
```{r calculate GDP}
calcGDP(gapminder_data, year=1967, country="Denmark")
calcGDP(gapminder_data, year=1977, country="Denmark")
calcGDP(gapminder_data, year=1987, country="Denmark")
calcGDP(gapminder_data, year=1997, country="Denmark")
calcGDP(gapminder_data, year=2007, country="Denmark")
calcGDP(gapminder_data, year=2017, country="Denmark")

```
The GDP was 77116977700 in 1967, 103920280028 in 1977, 128771236166 in 1987, 157476118456 in 1997, and 192906627081 in 2007. It isn’t possible to calculate the GDP from 2017, because there isn’t any data from this year in the data set.  

## Task 2

I need to write a script that loop over each country in the `gapminder` dataset, tests whether the country starts with a ‘B’, and prints out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r loop}
#creating a variable containing only the countries that starts with “B"
countries_with_B <- grep("^B", unique(gapminder$country), value=TRUE)
countries_with_B

#Creating two threshold values which is 50 and 70.
thresholdvalue1 <- 50
thresholdvalue2 <- 70

#creating the function

for (i in countries_with_B) {
  tmp <- mean(gapminder[gapminder$country == i, "lifeExp"], na.rm = TRUE)
  if (tmp < thresholdvalue1) {
    cat("Average Life Expectancy in", i, "is less than", thresholdvalue1, "\n")
  } else if (tmp > thresholdvalue2) {
    cat("Average Life Expectancy in", i, "is greater than", thresholdvalue2, "\n")
  } else {
    cat("Average Life Expectancy in", i, "is between", thresholdvalue1, "and", thresholdvalue2, "\n")
  } 
  rm(tmp)
} 

```

